---
title: "Cleaning_raw_biometrics"
author: "Tiff Stephens"
date: "11/30/2017"
output: html_document
---


```{r Libraries, echo = FALSE, include=FALSE}
library(dplyr); library(tidyr); library(ggplot2)
theme_set(theme_classic())
```

Import data
```{r}
# datachunk creator: TAS

eg.bio <- read.csv("https://raw.githubusercontent.com/APECS-ak/seagrass_2017/master/RAW_seagrass_biometrics_2017.csv", stringsAsFactors = FALSE, header = TRUE)
```

# Checking the data
```{r}
# datachunk creator: WR

str(eg.bio)
range(eg.bio$rhi_length, na.rm = TRUE)
range(eg.bio$node1, na.rm = TRUE)
hist(eg.bio$node2)
hist(eg.bio$leaf_width1)
```

Change numerics to factors
```{r}
# datachunk creator: WR

eg.bio$quadrat <- as.factor(eg.bio$quadrat)
eg.bio$plant <- as.factor(eg.bio$plant)
```



# Cleaning for analysis

Calculate dry weights of all flora and fauna; remove pad + foil data uneccessary for analysis
```{r}
# datachunk creator: TAS


# piped calculations for all
eg.bio = eg.bio %>% 
  mutate(epiphyte_mass_dw = pad_epiphyte_mass_g - pad_mass_g) %>%
  mutate(rhi_mass_dw = rhi_foil_dw_g - rhi_foil) %>%
  mutate(shoot_mass_dw = shoot_foil_dw - shoot_foil) %>%
  mutate(xs_shoot_mass_dw = xs_shoot_foil_dw - xs_shoot_foil) %>%
  mutate(xs_pad_mass_dw = xs_epiphyte_pad_mass_g - xs_pad_mass_g) %>%
  mutate(iso_mass_dw = iso_foil_dw - iso_foil) %>%
  mutate(gamm_amph_mass_dw = gamm_amph_foil_dw - gamm_amph_foil) %>%
  mutate(caprel_mass_dw = caprel_foil_dw - caprel_foil) %>%
  mutate(limpet_mass_dw = limpet_foil_dw - limpet_foil) %>%
  mutate(gastropod_mass_dw = gastropod_foil_dw - gastropod_foil) %>%
  mutate(crab_mass_dw = crab_foil_dw - crab_foil) 

# check new column names
colnames(eg.bio)

# remove pad and foil weights from dataframe 
eg.bio = eg.bio %>% 
  select(-pad_epiphyte_mass_g, -pad_mass_g, -xs_epiphyte_pad_mass_g, -xs_pad_mass_g) %>% #pads
  select(-rhi_foil_dw_g, -rhi_foil, -shoot_foil, -shoot_foil_dw, -xs_shoot_foil_dw, -xs_shoot_foil) %>% #seagrass
  select(-iso_foil_dw, -iso_foil, -gamm_amph_foil_dw, -gamm_amph_foil, -caprel_foil_dw, -caprel_foil, -crab_foil_dw, -crab_foil) %>% #crustaceans
  select(-limpet_foil_dw, -limpet_foil, -gastropod_foil_dw, -gastropod_foil) #molluscs

# check new column names
colnames(eg.bio)
```


Calculate leaf areas for each blade measured
```{r, echo=FALSE}
# datachunk creator: TAS


# calculate the leaf area (cm^2) per blade per plant
eg.bio = eg.bio %>% 
  mutate(leaf_area1 = leaf_length1 * leaf_width1, na.rm = TRUE) %>%
  mutate(leaf_area2 = leaf_length2 * leaf_width2, na.rm = TRUE) %>%
  mutate(leaf_area3 = leaf_length3 * leaf_width3, na.rm = TRUE) %>%
  mutate(leaf_area4 = leaf_length4 * leaf_width4, na.rm = TRUE) %>%
  mutate(leaf_area5 = leaf_length5 * leaf_width5, na.rm = TRUE) %>%
  mutate(leaf_area6 = leaf_length6 * leaf_width6, na.rm = TRUE) %>%
  mutate(leaf_area7 = leaf_length7 * leaf_width7, na.rm = TRUE) %>%
  mutate(leaf_area8 = leaf_length8 * leaf_width8, na.rm = TRUE) %>%
  mutate(leaf_area9 = leaf_length9 * leaf_width9, na.rm = TRUE) %>%
  mutate(leaf_area10 = leaf_length10 * leaf_width10, na.rm = TRUE)

# check new column names
colnames(eg.bio)
```

Calculate total leaf area for each plant
```{r}
# datachunk creator: TAS


# calculate total leaf area per plant (there is no biomass data per blade so need to combine all blades)
eg.bio$tot_leaf_area <- rowSums(eg.bio[,c("leaf_area1", "leaf_area2", "leaf_area3", "leaf_area4", "leaf_area5", "leaf_area6", "leaf_area7", "leaf_area8", "leaf_area9", "leaf_area10")], na.rm = TRUE)

# check new column names
colnames(eg.bio)
```

Calculate mean canopy height (all blades from each plant included)
```{r}
# datachunk creator: TAS


# mean height of all blades
eg.bio = mutate(eg.bio, mean_leaf_length = rowMeans(select(eg.bio, starts_with("leaf_length")), na.rm = TRUE))

# check new column names
colnames(eg.bio)
```

Determine max canopy height per plant (only tallest blade from each plant included)
```{r, warning = FALSE, message = FALSE}
# datachunk creator: TAS

eg.bio$max_canopy_height = apply(eg.bio[, 12:21], 1, max, na.rm = TRUE)

colnames(eg.bio)
```

```{r}
write.csv(eg.bio,'CLEAN_seagrass_biometrics_2017.csv')
```



































